---
title: "Discrete distributions "
format: 
  html:
    self-contained: true
    table-of-contents: true
    number-sections: true
editor: visual
execute: 
  eval: true
  warning: false
  message: false
---

```{r setup, echo = FALSE}
JuliaCall::julia_setup(JULIA_HOME = "/Applications/Julia-1.8.app/Contents/Resources/julia/bin/")
knitr::opts_chunk$set(engine.path = list(
  python = "/Library/Frameworks/Python.framework/Versions/3.10/bin/python3.10",
  julia = "/Applications/Julia-1.8.app/Contents/Resources/julia/bin/"
))
```

## Exercises from lectures (2023.02.28)

::: panel-tabset
## R

Exercise 1: generate 10 pseudo-random numbers from Bernoulli(0.7)

```{r}
set.seed(20230228)
rbinom(n=10,size=1,prob=0.7)
```

Exercise 2: generate 10 pseudo-random numbers from Binomial(50,0.7)

```{r}
set.seed(20230228)
rbinom(n=10,size=50,prob=0.7)
```

Exercise 3: generate 10 pseudo-random numbers from Poisson(1.5)

```{r}
set.seed(20230228)
rpois(n=10,lambda=1.5)
rpois(10,1.5)
```

## Python

Load modules

```{python}
import scipy.stats as st
import numpy as np
```

Exercise 1: generate 10 pseudo-random numbers from Bernoulli(0.7)

```{python}
np.random.seed(20230228)
st.bernoulli(0.7).rvs(10)
```

Exercise 2: generate 10 pseudo-random numbers from Binomial(50,0.7)

```{python}
np.random.seed(20230228)
st.binom(50,0.7).rvs(10)
```

Exercise 3: generate 10 pseudo-random numbers from Poisson(1.5)

```{python}
np.random.seed(20230228)
st.poisson(1.5).rvs(10)
```

## Julia

Load packages

```{julia}
using Distributions
using Random
```

Exercise 1: generate 10 pseudo-random numbers from Bernoulli(0.7)

```{julia}
Random.seed!(20230228);
rand(Bernoulli(0.7),10)
```

Exercise 2: generate 10 pseudo-random numbers from Binomial(50,0.7)

```{julia}
Random.seed!(20230228);
rand(Binomial(50,0.7),10)
```

Exercise 3: generate 10 pseudo-random numbers from Poisson(1.5)

```{julia}
Random.seed!(20230228);
rand(Poisson(1.5),10)
```
:::

## Software versions {.unnumbered}

``` R
R version 4.2.1 (2022-06-23)
Platform: x86_64-apple-darwin17.0 (64-bit)
Running under: macOS Big Sur 11.7.3
```

``` python
Python 3.10.7 (v3.10.7:6cc6b13308, Sep  5 2022, 14:02:52) [Clang 13.0.0 (clang-1300.0.29.30)] on darwin
Type "help", "copyright", "credits" or "license" for more information.
```

``` julia
Julia Version 1.8.4
Commit 00177ebc4fc (2022-12-23 21:32 UTC)
```
