---
title: "Generalized linear regression: logistic regression"
format: 
  html:
    self-contained: true
    table-of-contents: true
    number-sections: true
    df-print: kable
editor: visual
execute: 
  eval: true
  warning: false
  message: false
---


```{r setup, echo = FALSE}
JuliaCall::julia_setup(JULIA_HOME = "/Applications/Julia-1.9.app/Contents/Resources/julia/bin/")
knitr::opts_chunk$set(engine.path = list(
  python = "/Library/Frameworks/Python.framework/Versions/3.10/bin/python3.10",
  julia = "/Applications/Julia-1.9.app/Contents/Resources/julia/bin/"
))
```


## Setup

::: panel-tabset
## R

Install packages if necessary.

```{r, eval = FALSE}
install.packages("marginaleffects")
install.packages("modelsummary")
install.packages("car")
install.packages("broom")
```

```{r}
library(marginaleffects)
library(modelsummary)
library(car)
library(broom)
```

Read in the dataset for the lecture and specify column classes (types).

```{r}
df <- read.csv("../data/polish-jvs.csv", colClasses = c("character", "factor", rep("character", 4), "numeric"))
head(df)
```

## Python

Load modules

```{bash, eval = F}
!pip install statsmodels 
```

```{python}
import pandas as pd
import numpy as np
import statsmodels.api as sm
import statsmodels.formula.api as smf
```

Read in the dataset for the lecture and specify column classes (types).

```{python}
df = pd.read_csv("../data/polish-jvs.csv", dtype={"id": np.int64, "woj":str, "public":str,"size": str, "nace_division": str, "nace": str})
df.head()
```

## Julia

```{julia, eval = F}
using Pkg
Pkg.add(["Effects", "StatsBase", "CSV", "CategoricalArrays", "GLM"])
```

Load packages

```{julia}
using Effects
using StatsBase
using CSV
using CategoricalArrays
using GLM
using Statistics
using DataFrames
```

Read in the dataset for the lecture and specify column classes (types).

```{julia}
df=CSV.read("../data/polish-jvs.csv", DataFrame, 
            types = Dict(:id => Int64, :woj=> String, :public=> String, 
                         :size => String, :nace_division => String, :nace => String));
first(df, 5)
```

:::

## Example 1 -- individual data

Create new variable 

$$
\text{has_vacancies} = 
\begin{cases}
1 & \text{if has at least one vacancy},\\
0 & \text{otherwise}
\end{cases}
$$

::: panel-tabset
## R

```{r}
df$has_vacancy <- df$vacancies > 0
df$size <- factor(df$size, c("Large", "Medium", "Small"))
```

```{r}
model1 <- glm(formula = has_vacancy ~ size + public, 
              data = df,
              family = binomial())
summary(model1)
```

```{r}
modelsummary(model1, exponentiate = T, conf_level = 0.05, 
             statistic  = NULL,
             estimate  = "{estimate} [{conf.low}, {conf.high}]")
```

```{r}
car::Anova(model1, type = 2) |> broom::tidy() |> as.data.frame()
```

## Python

## Julia

:::


## Example 2 -- aggregated data


::: panel-tabset
## R

```{r}
df_aggregated <- aggregate(vacancies ~ has_vacancy + public + size, data = df, FUN = length) |>
  reshape(data = _, direction = "wide", idvar = c("public", "size"), timevar = "has_vacancy", sep = "_")

df_aggregated
```

```{r}
model2 <- glm(formula = cbind(vacancies_TRUE, vacancies_FALSE) ~ size + public,
              data = df_aggregated, 
              family = binomial())
summary(model2)
```
```{r}
car::Anova(model2, type = 2) |> broom::tidy() |> as.data.frame()
```

## Python

## Julia

:::

